---
title: "Symbolic Computation in R"
author: "Carlos Calvo Hernandez"
date: "5/07/2019"
output: 
  html_document:
    toc: TRUE
    toc_float: TRUE
---

```{r setup, include=FALSE}
# Use `sudo R CMD javareconf` and restart RStudio if the folder to the jdk is not found
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE, cache = TRUE)
```

# Using `rSymPy` 

We will start by using the `rSymPy` package that uses Python and Java. SymPy is a Python library and the R package connects to that library.

## Installing SymPy

SymPy can be installed on virtually any computer with Python 2.7 or above. SymPy does require *mpmath* Python library to be installed first. The recommended method of installation is through **Anaconda**, which includes *mpmath*, as well as several other useful libraries.

### What is Anaconda?

Anaconda is a free Python distribution from Continuum Analytics that includes *SymPy, Matplotlib, IPython, NumPy*, and many more useful packages for scientific computing. This is recommended because many nice features of *SymPy* are only enabled when certain libraries are installed.

### First, installing Python through Anaconda

Go to the [Anaconda Distribution](https://www.anaconda.com/distribution/) webpage and download the Python 3.7[^1] Graphical Installer version for your OS. Run the installer and follow the directions. Aaaaaaand... you're done! Really, that's it. You are now the proud owner of a fully functional Python distribution that can be accessed through the Anaconda software.

[^1]: I recomend 3.7 over 2.7 since Python 2.x is losing official support starting 2021.

### Finally, installing SymPy with *conda*

Now that you have Anaconda, we need to make sure all libraries we need are installed. Go to a terminal window and type:

> conda install mpmath

If *mpmath* is already installed, just update it:

> conda update mpmath

Install *sympy*

> conda install sympy

That's it! Ready to use!

## `rSymPy``

Now install the R package and let's use it.

```{r, eval=FALSE}
install.packages("rSymPy")
```

```{r rsympy}
library(rSymPy)
```

We can use the `var()` function to create a SymPy variable. And subsequently use that variable with any mathematical operators. 

```{r}
x <- Var("x")
x+x
```

```{r}
x*x/x
```

We can operate with multiple variables:

```{r}
y <- Var("x**3")
x/y
```

We can also just use the generic function `sympy` to create the variables. `sympy` can be used to create expressions or variables.

```{r}
z <- "2.5*x**2"
z + y
```

`sympy` also uses the method `evalf()` to evaluate a numerical expression into a floating point number:
```{r}
sympy("sqrt(8)")
sympy("sqrt(8).evalf()")
sympy("sqrt(8).evalf(50)")
```

The index `n` inside `evalf(n)` is the number of decimal places to be displayed
```{r}
sympy("pi.evalf(120)")
```

We can assign names to `sympy` expressions within the function:

```{r}
sympy("one = cos(1)**2 + sin(1)**2")
sympy("(one - 1)")
sympy("(one - 1).evalf()") # rounding errors from floating number arithmetics
```

When not interested in the floating number rounding errors we can use the option `chop=TRUE` in `evalf()`

```{r}
sympy("(one - 1).evalf(chop=True)")
```

`sympy` also handles equations with the method `Eq()`:

```{r}
sympy("Eq(x**2+2*x+1,(x+1)**2)")
```

```{r}
sympy("a = x**2+2*x+1")
sympy("b = (x+1)**2")
```

We can also symplify equations

```{r}
"0" == sympy("simplify(a-b)")
```

```{r}
sympy("simplify((x**3 + x**2 - x - 1)/(x**2 + 2*x + 1))")
```

We can expand and factorize accordingly

```{r}
sympy("(x + 2)*(x - 3)")
sympy("expand((x + 2)*(x - 3))")
```
```{r}
sympy("factor(x**3 - x**2 + x - 1)")
```

`collect()` organizes equations around a given variable:

```{r}
y <- Var("y")
z <- Var("z")
sympy("collect(x*y + x - 3 + 2*x**2 - z*x**2 + x**3, x)")
```

```{r}
sympy("(x*y**2 - 2*x*y*z + x*z**2 + y**2 - 2*y*z + z**2)/(x**2 - 1)")
sympy("cancel((x*y**2 - 2*x*y*z + x*z**2 + y**2 - 2*y*z + z**2)/(x**2 - 1))")
```

```{r}
sympy("expand_func(gamma(x + 3))")
```

We can create variables that have a persistent state, since objects in SymPy are immutable.

```{r}
sympy("y = x*x")
```

This persistent variable can then be used in other expression. E.g. in matrices.

```{r}
sympy("A = Matrix([[1,x], [y,1]])")
sympy("A**2")
```

We can substitute variables with other variables or values.

```{r}
sympy("B = A.subs(x,1.1)")
sympy("B**2")
```

```{r}
# more replacement, a subexpression by another:
sympy("expr = sin(2*x) + cos(2*x)")

sympy("expr.subs(sin(2*x), 2*sin(x)*cos(x))")

sympy("expr.subs(x,pi/2)")
```

There are matrix operations with `inv()` and `det()`

```{r}
a1 <- Var("a1")
a2 <- Var("a2")
a3 <- Var("a3")
a4 <- Var("a4")

A <- Matrix(List(a1, a2), List(a3, a4))
#define inverse and determinant
Inv <- function(x) Sym("(", x, ").inv()")
Det <- function(x) Sym("(", x, ").det()")

A
```
```{r}
cat(A,"\n")
```

```{r}
Inv(A)
```

```{r}
Det(A)
```

Using the `exp()` function:

```{r}
Exp <- function(x) Sym("exp(", x, ")") 
Exp(-x) * Exp(x)
```

Solve equations with `solve()``

```{r}
sympy("solve(x**2 - 2, x)")
```

Calculate limits with `limit()`

```{r}
sympy("limit(1/x, x, oo)")
sympy("limit(1/x, x, 0)")
```

`integrate()` handles definite and indefinite integrals:

```{r}
sympy("integrate(exp(-x))")                   # indefinite integral
```
```{r}
sympy("integrate(exp(-x*y),x)")               # indefinite integral
```
```{r}
sympy("integrate(exp(-x), (x, 0, oo))")       # definite integral
```


Integration is possible in base R, but the syntax is not straightforward:

```{r}
integrate( function(x) exp(-x), 0, Inf)
```

`diff()` handles derivatives

```{r}
sympy("diff(sin(2*x), x, 1)")       # first derivative
sympy("diff(sin(2*x), x, 2)")       # second derivative

sympy("diff(sin(2*x), x, 3)")       # third  derivative


sympy("diff(exp(x*y*z), x, z, 3)")  # d^4/dxdz^3
```

Differentiation is also possible in base R:

```{r}
D( expression(sin(2*x)), "x" )
```

`series()` handles all series

```{r}
sympy("(1/cos(x)).series(x, 0, 10)")  # taylor expansion

sympy("exp(x).series(x, 0, 5)")       # taylor expansion

sympy("exp(x).series(x, 0, 5).removeO()")
```

`eigenvals()` calculates the eigenvalues of a matrix:

```{r}
sympy("Matrix([[1, 2], [2, 2]]).eigenvals()")
```

And, finally, we can output everything directly to LaTex with `latex()`

```{r}
sympy("latex(Integral(cos(x)**2, (x, 0, pi)))")
```

This can be used within an RMarkdown document like this $\int_{0}^{\pi} \operatorname{cos}^{2}\left(x\right)\,dx = \frac{1}{2} \pi$

## References

1. `rSymPy` package reference manual: <http://cran.r-project.org/web/packages/rSymPy/rSymPy.pdf>
2. SymPy Tutorials: <http://docs.sympy.org/latest/tutorial/>
